---
title: "Lichess Elite April 2021"
author: "Gary Lu"
date: "5/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(bigchess)
library(knitr)
```

### Analysis of April 2021 Chess games from the Lichess Elite Database at https://database.nikonoel.fr/
I don't play much chess but I have gotten an interest in watching games recently so I wanted to grab a small dataset and look at the trends


```{r load data, echo=FALSE, include=FALSE}

## Converts pgn to csv to load faster when opening again

#  chess_games_apr_21 <- read.pgn("lichess_elite_2021-04.pgn",
#                                   add.tags = c("Event", "LichessURL", "Date", "Round", "White", "Black", "WhiteElo", "BlackElo",
#                                                "ECO", "Opening", "TimeControl", "UTCDate", "UTCTime", "Termination"),
#                                   n.moves = T,
#                                   last.move = F,
#                                   stat.moves = F
#                                  )
# 
# write_csv(chess_games_apr_21, "C:\\Users\\thegl\\Documents\\R\\lichess_elite_apr21.csv")


df <- read_csv("lichess_elite_apr21.csv")
df <- df %>% 
  filter(complete.movetext == TRUE) %>% 
  unite(M10, 19:38, sep = " ", remove = FALSE) # Unites the first 10 moves of each side into 1 column but keeps individual move columns

```

```{r}
# Helper Functions

row_search <- function(df, col, string){
  df %>% 
    filter(str_detect(df[[col]], str_c("^", string)))
}
```

### What time controls are played the most:
```{r}
group_count(df, "TimeControl") %>% 
  head(10)
```
There is a heavy favor towards 3 minute games

### Breakdown of game results

```{r, echo = FALSE}
# Winrate of games
results <- c("0-1", "1/2-1/2", "1-0")

game_results <- 
  df %>% 
  group_by(Result) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2), game = "") %>% 
  arrange(factor(Result, level = results)) %>% 
  mutate(Result = factor(Result, levels = results))


   ggplot(game_results)+
   geom_col(aes(x = game, y = percent, fill = Result), width = .1)+
   coord_flip()+
     labs(x = "",
          title = "Game Results")+
     theme_light()
```
White has a slightly better winrate than black here


### The most popular openings for each side

```{r, echo=FALSE}
popular_openings <- 
  df %>% 
  group_count("W1") %>% 
  mutate(percent = n / sum(n) * 100) %>%
  head(10)

ggplot(data = popular_openings)+
  geom_col(aes(x = reorder(W1, percent), y = percent, fill = W1))+
  coord_flip()+
  labs(title = "Most popular openings for white")

```

The ten most popular openings for white. e4 and d4 follow the concept of putting pieces in the center

```{r, echo = FALSE}
df %>% 
  filter(W1 %in% popular_openings$W1) %>% 
  group_by(W1, Result) %>% 
  summarise(n = n()) %>% 
  mutate(Result = factor(Result, levels = results)) %>% 
  arrange(factor(Result, level = results)) %>%
  ggplot()+
  geom_col(aes(x = reorder(W1, n), y = n, fill = Result), position = "fill")+
  coord_flip()+
  labs(x = "",
       y = "percent",
       title = "Winrate of popular openings")
```

e4 being the most popular 1st move does not have the highest win rate for white
```{r}
df %>% 
  filter(W1 %in% popular_openings$W1) %>% 
  group_by(W1) %>% 
  summarise(avgNMoves = mean(NMoves))
```

No matter which of the popular openings is played, games last roughly the same number of moves

### Getting into specific opening varitions:

```{r, warning = FALSE}
dfm1 <- 
  df %>% 
  unite(M1, W1, B1, sep = " ")


popularm1 <-   
  dfm1 %>% 
  group_count(c("M1")) %>% 
  head(10)
   
  ggplot(popularm1)+
  geom_col(aes(x = reorder(M1, n), y = n, fill = M1))+
  coord_flip()+
  labs(x = "First moves",
       y = "",
       title = "Most popular opening moves")
```


When playing as black, people avoid playing symmetrical openings (although they are the 2nd most popular choice)
 
### Let's look at the most popular continuations to each opening

```{r}
top5 <- function(move1){
  dfm1 %>% 
    filter(M1 == move1) %>% 
    group_by(M1, Opening) %>% 
    summarise(n = n()) %>% 
    arrange(desc(n)) %>% 
    head(5)
}

map(popularm1$M1, top5)
```

The top 5 continuations for each opening. Now that we know names of popular openings we can look into each individual one

```{r}

ruylopez <- row_search(df, "Opening", "Ruy Lopez")

ruylopez %>% group_count("Result") %>% 
             mutate(percent = n / sum(n) * 100)

sicilian <- row_search(df, "Opening", "Sicilian") 

sicilian %>% group_count("Result") %>% 
              mutate(percent = n / sum(n) * 100)

```

Results of the Ruy Lopez, one the first openings I learned of although much less popular compared to the Sicilian Defense. The use of the helper functions helps to narrow down what kind of games I want to look at

The moves for the Ruy Lopez are e4 e5 Nf3 Nc6 Bg5 so we start with blacks 3rd move
```{r}
rlresults <- group_count(ruylopez, c("B3", "Result")) %>% 
              arrange(desc(n))

rlresults$B3 <- factor(rlresults$B3)
rlresults$Result <- factor(rlresults$Result, levels = c("0-1", "1/2-1/2", "1-0"))

ggplot(rlresults %>% head(30))+
  geom_col(aes(x = fct_reorder(B3, n), y = n, fill = Result))+
  coord_flip()
  
```

For the most part, Ruy Lopez games tend to go the same way for the first 5 moves