---
title: "Lichess Elite March 2021"
author: "glu"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(bigchess)
library(knitr)
```

### Analysis of March 2021 Chess games from the Liches Elite Database at https://database.nikonoel.fr/
### Since this covers only games on lichess with ratings 2200-2400 (excluding bullet games) the sample size is smaller than the general population

```{r load data, echo=FALSE}
chess_games_march_21 <- read.pgn("lichess_elite_2021-03.pgn",
                                 add.tags = c("Event", "LichessURL", "Date", "Round", "White", "Black", "WhiteElo", "BlackElo",
                                              "ECO", "Opening", "TimeControl", "UTCDate", "UTCTime", "Termination"),
                                 n.moves = F,
                                 last.move = F,
                                 stat.moves = F
)

df <- chess_games_march_21 %>% filter(complete.movetext == TRUE) # Filtering out NAs
```

## Breakdown of game results
```{r}
# Winrate of games
df %>% 
  group_by(Result) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>%
  arrange(desc(percent))
```

### White winrate seems to be slightly better than black if you don't factor in draws

## The most popular openings for each side

```{r, echo=FALSE}
count_openings <- df %>% 
  group_by(W1) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>%
  arrange(desc(n)) %>% head(10)

ggplot(count_openings)+
  geom_col(data = count_openings, aes(x = reorder(W1, percent), y = percent, fill = W1))+
  coord_flip()+
  labs(title = "Most popular openings for white")
```

```{r}
df %>%
  filter(Result == "1-0") %>% 
  group_by(W1) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>%  head(10) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(W1, percent), y = percent, fill = W1))+
  coord_flip()+
  labs(title = "Win percentage for white openings",
       x = "")
```

## Most games that result in a 1-0 are from e4, which correlates to it's popularity

## Exploring responses to common openings:

```{r, echo = False}
e4b1 <- df %>% 
  filter(!is.na(B1) & W1 == "e4") %>% 
  group_by(B1)

e4b1 %>% 
  summarise(n=n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% head(10) %>% 
ggplot()+
  geom_col(aes(x = reorder(B1, percent), y = percent, fill = B1))+
  coord_flip()+
  labs(title = "Popular Black Responses to E4",
       x = "")
```

## c5 (Sicilian Defense) is by far the most popular e4 reponse here

```{r, echo = False}
common_b1 <- c("c5", "e5", "e6", "c6", "d5", "d6", "g6", "Nf6", "Nc6", "b6")
e4b1 %>% 
  filter(Result == "0-1" | Result == "1/2-1/2") %>% 
  filter(B1 %in% common_b1)%>% 
  group_by(B1, Result) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(percent)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(B1, n), y = percent, fill = Result), position = "dodge")+
  coord_flip()+
  labs(title = "Result breakdown of Black Responses to e4",
       x = "")
```
### It seems like c5, despite being the most popular response, does not have the highest win rate or draw rate (again not counting sample size). Since it is the most popular response, more people would learn how to play against it

```{r}
b1wr <- 
e4b1 %>% 
  filter(B1 %in% common_b1)%>% 
  group_by(B1, Result) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  filter(Result == "0-1" | Result == "1/2-1/2") %>% 
  arrange(desc(percent))

b1wr %>% 
  filter(Result != "1-0") %>% 
  group_by(B1) %>% 
  summarise(n = sum(n), percent = sum(percent)) %>% 
  arrange(desc(percent)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(B1, percent), y = percent, fill = B1))+
  coord_flip()+
  labs(title = "Non-Loss games for black",
       x = "")
```
### Factoring in draws, c5 still isn't the top. However, the sample sizes are different and the results are not too far from each other.

## Repating the same process for d4:
```{r, echo = False}
d4b1 <- df %>% 
  filter(!is.na(B1) & W1 == "d4") %>% 
  group_by(B1)

d4b1 %>% 
  summarise(n=n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% head(10) %>% 
ggplot()+
  geom_col(aes(x = reorder(B1, percent), y = percent, fill = B1))+
  coord_flip()+
  labs(title = "Popular Black Responses to d4")
```

```{r, echo = FALSE}
common_b1d4 <- c("Nf6", "d5", "e6", "g6", "c5", "d6", "f5", "c6", "e5", "Nc6")
d4b1 %>% 
  filter(B1 %in% common_b1d4)%>% 
  filter(Result == "0-1" | Result == "1/2-1/2") %>% 
  group_by(B1, Result) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(percent)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(B1, n), y = percent, fill = Result), position = "dodge")+
  coord_flip()+
  labs(title = "Result breakdown of Black Responses to d4")
```

## Nf6 is the most popular d4 response by a longshot. It is similar to the e4 response chart in the most popular response is nearly double the second most popular response, but it falls into the same trend where it is not the highest percent for a 0-1 or a 1/2 - 1/2 result

## Getting into specific openings:

# Which opening variation is the most popular?

```{r, echo = FALSE}
e4openings <- 
df %>% 
  filter(W1 == "e4") %>% 
  group_by(Opening) %>% 
  summarize(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% 
  head(10)

d4openings <- 
df %>% 
  filter(W1 == "d4") %>% 
  group_by(Opening) %>% 
  summarize(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% 
  head(10)
```

# Results break down of most popular openings
```{r}
df %>% 
  filter(Opening %in% e4openings$Opening) %>% 
  group_by(Opening, Result) %>% 
  summarize(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(Opening, n), y = percent, fill = Result), position = "dodge")+
  coord_flip()+
  labs(title = "Results of e4 Opening Variations")
  

df %>% 
  filter(Opening %in% d4openings$Opening) %>% 
  group_by(Opening, Result) %>% 
  summarize(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% 
  ggplot()+
  geom_col(aes(x = reorder(Opening, n), y = percent, fill = Result), position = "dodge")+
  coord_flip()+
  labs(title = "Results of d4 Opening Variations")


```

## Among the popular e4 openings, white tends to win more except in the Najdorf Sicilian. In all cases with d4, white is better