---
title: "Looking at the NY Times public dataset for Covid-19 cases"
author: "Gary Lu"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
```

```{r, raw_data, include = FALSE}
covid_index_states <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
covid_index_us <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us.csv")
today <- tail(covid_index_states,1)$date # Setting "today" to the latest date
```

## Reporting information on data from NYTimes public dataset 

All data was retrieved from <https://github.com/nytimes/covid-19-data>

It includes the datasets:

<https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv>

<https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv>

<https://raw.githubusercontent.com/nytimes/covid-19-data/master/us.csv>

## Most recent data reported

## Total number of reports in the US
```{r us_index, echo = FALSE}
us_totals <- covid_index_us %>% 
             mutate(delta_cases = 
                 coalesce(cases - lag(cases), 0)) %>% 
             mutate(delta_deaths = 
                 coalesce(deaths - lag(deaths), 0)) 

tail(us_totals) %>% kable(caption = "Last week's numbers")
```


## From the public github:

*May 7, 2020*

*At the beginning of the Covid-19 epidemic in the United States, most health departments and the Centers for Disease Control only considered a case to be “confirmed” as Covid-19 with a positive confirmatory laboratory test result for SARS-CoV-2 RNA. Recently, however, more and more states and localities have started to include “probable” cases and/or deaths in their counts — in some places reporting only a combined “confirmed” and “probable” total.*

*Because of these developments, we have changed our Covid-19 dataset to start including “probable” Covid-19 cases and deaths when they are available. It is no longer possible to report a comprehensive “confirmed-only” dataset.*

*This change could cause discontinuities in the data when a locality begins to report “probable” cases and deaths. We are working to acquire the historical data from those places if and when they make it available to continually improve the quality of the dataset.*

Definition of probable:

*On April 5, the Council of State and Territorial Epidemiologists issued a recommendation that states include both confirmed cases, based on laboratory testing, and probable cases, based on specific criteria for symptoms and exposure. The C.D.C. adopted these definitions and began reporting probable cases on April 14 when New York City announced an additional 3,700 deaths presumed to have been caused by Covid-19.*

*The new criteria define a “probable” Covid-19 case or death based on a combination of evidence from clinical, epidemiologic or serological testing, or from vital records, but without a confirmatory laboratory RNA test.*

**With this in mind, it might not be surprising so see some odd findings in the dataset**

```{r, echo = FALSE, cache = TRUE, dependson="raw_data"}
us_today <- tail(covid_index_us, 1)

state_totals <- covid_index_states %>% 
                group_by(state) %>%               
                select(date, state, cases, deaths) %>%
                mutate(delta_cases = 
                        coalesce(cases - lag(cases, order_by = state), 0)) %>% 
                mutate(delta_deaths = 
                        coalesce(deaths - lag(deaths, order_by = state), 0))
```

## Checking if there are cases where there is a negative delta (cases reported one day are less than the previous)
```{r, echo = FALSE, cache = TRUE, dependson="raw_data"}
    us_totals %>% 
        filter(delta_cases < 0 | delta_deaths < 0) %>% 
        kable(caption = "From the us dataset")

    head(state_totals %>% 
        filter(delta_cases < 0 | delta_deaths < 0) %>% 
        select(date, state, delta_cases, delta_deaths))%>%
        kable(caption = "From the states dataset")
```

In the states dataset there are quite a number of instances where either the delta cases or delta deaths have been negative but the us dataset does not contain any negative deltas

### Adjusting the data to exclude any negative deltas
```{r, echo = FALSE, cache = TRUE, dependson="raw_data"}
states2 <- state_totals %>% 
      filter(delta_cases >= 0 & delta_deaths >= 0)
```

## Comparing statistical summaries with/without negative deltas
```{r, echo = FALSE}
summarise(state_totals, avg_new_cases = mean(delta_cases), avg_new_deaths = mean(delta_deaths)) %>% kable(caption = "With negative deltas")
summarise(states2, avg_new_cases = mean(delta_cases), avg_new_deaths = mean(delta_deaths)) %>% kable(caption = "Without negative deltas")
```
Calculating the mean of both has slightly different results when you take out negative deltas

## Comparing each state's proportion of total number of US cases

```{r, echo = FALSE, cache = TRUE, dependson="raw_data"}
total_index <- us_totals %>% 
                left_join (states2, by = "date") %>% 
                mutate(prop_cases = coalesce(cases.y / cases.x * 100, 0),
                       prop_deaths = coalesce(deaths.y / deaths.x * 100, 0))

today_index <- total_index %>% filter(date == today)
today_index %>% 
  arrange(desc(prop_cases)) %>% 
  select(date, state, prop_cases, prop_deaths) %>% 
  kable()
```

## Each state's proportion of total covid cases:
```{r, echo = FALSE, warning = FALSE}
us_map <- map_data("state")
today_index$state <- tolower(today_index$state)
today_index$prop_cases <- round(today_index$prop_cases, 2)


ggplot()+
  geom_map(data = today_index, map = us_map,
           aes(fill = prop_cases, map_id = state))+
   geom_polygon(data = us_map, aes(x = long, y = lat, group = group), color = "black", fill = NA)+
  coord_quickmap()+
  labs(x = "",
       y = "",
       title = "State percentage of total COVID cases",
       subtitle = "Based on data from NYTimes",
       fill = "Percentage of COVID cases")+
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()
    )+
  scale_fill_distiller(palette = 10, direction = 1)
```

```{r, echo = FALSE, warning = FALSE}
state_mean_cases <- summarise(state_totals, avg_new_cases = mean(delta_cases))
state_mean_cases$state <- tolower(state_mean_cases$state)

state_mean_cases %>% 
  ggplot()+
    geom_map(map = us_map,
            aes(fill = avg_new_cases, map_id = state))+
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), color = "black", fill = NA)+
  coord_quickmap()+
  scale_fill_distiller(palette = 8, direction = 1)+
  labs(x = "",
       y = "",
       title = "Average new cases per day",
       subtitle = "Based on data from NYTimes",
       fill = "Average New Cases per day")+
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()
    )
```
