---
title: "Analysis of the Wordle word list"
author: "Gary Lu"
date: "3/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

### This may contain **SPOILERS** for Wordle.

#### The wordlist was extracted from the developer tools in a web browser, however it also contains **_SPOILERS_** for future days so I had to be careful when looking at this

#### _This information is not meant to be taken seriously as you should be playing Wordle the way you want to._

I will take a look a common letters, as well as letter patterns 

```{r, echo = FALSE, message = FALSE}
wordlist <- read_delim("wordlist.txt", delim = ",")

wordlist <- tibble("word" = colnames(wordlist))
wordlist$word <- toupper(wordlist$word)

wordlist <- wordlist %>% mutate(L1 = str_sub(wordlist[[1]], 1, 1),
                    L2 = str_sub(wordlist[[1]], 2, 2),
                    L3 = str_sub(wordlist[[1]], 3, 3),
                    L4 = str_sub(wordlist[[1]], 4, 4),
                    L5 = str_sub(wordlist[[1]], 5, 5),
                    F3 = str_sub(wordlist[[1]], 1, 3),
                    M3 = str_sub(wordlist[[1]], 2, 4),
                    E3 = str_sub(wordlist[[1]], 3, 5))
```


```{r, echo=FALSE}
commonletters <- wordlist %>%
  select(contains("L")) %>% 
  pivot_longer(everything(), names_to = "pos", values_to = "letter") %>% 
  group_by(pos) %>% 
  count(letter) %>% 
  arrange(desc(n), .by_group = TRUE) %>% 
  slice_head(n = 5) %>% 
  ggplot(aes(x = fct_reorder(letter, -n), y = n, fill = pos))+
  geom_col(position = position_dodge())+
  labs(title = "Most common letters and their position",
       y = "",
       x = "Letter")

commonf3 <- wordlist %>% 
  select(contains("F3")) %>% 
  count(F3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(F3, -n), y = n, fill = F3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common first 3 letters",
       x = "",
       y = "")

commonm3 <- wordlist %>% 
  select(contains("M3")) %>% 
  count(M3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(M3, -n), y = n, fill = M3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common middle 3 letters",
       x = "",
       y = "")

commone3 <- wordlist %>% 
  select(contains("E3")) %>% 
  count(E3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(E3, -n), y = n, fill = E3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common last 3 letters",
       x = "",
       y = "")

commonletters
```
- S is the most common 5th letter probably because plural words
- A E and I are popular in at least 3 different positions
- I expected T and N to be more popular as anyone that has seen Wheel of Fortune knows how they are used

If were were to remove S from the 5th letter:

```{r}
wordlist2 <- wordlist %>% filter(L5 != "S")

nos_letters <- wordlist2 %>%
  select(contains("L")) %>% 
  pivot_longer(everything(), names_to = "pos", values_to = "letter") %>% 
  group_by(pos) %>% 
  count(letter) %>% 
  arrange(desc(n), .by_group = TRUE) %>% 
  slice_head(n = 5) %>% 
  ggplot(aes(x = fct_reorder(letter, -n), y = n, fill = pos))+
  geom_col(position = position_dodge())+
  labs(title = "Most common letters and their position",
       subtitle = "S removed from 5th letter",
       y = "",
       x = "Letter")

nos_f3 <- wordlist2 %>% 
  select(contains("F3")) %>% 
  count(F3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(F3, -n), y = n, fill = F3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common first 3 letters",
       subtitle = "S removed from 5th letter",
       x = "",
       y = "")

nos_m3 <- wordlist2 %>% 
  select(contains("M3")) %>% 
  count(M3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(M3, -n), y = n, fill = M3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common middle 3 letters",
       subtitle = "S removed from 5th letter",
       x = "",
       y = "")

nos_e3 <- wordlist2 %>% 
  select(contains("E3")) %>% 
  count(E3) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_reorder(E3, -n), y = n, fill = E3))+
  geom_col()+
  theme(legend.position = "")+
  labs(title = "Most common last 3 letters",
       subtitle = "S removed from 5th letter",
       x = "",
       y = "")


```

Side by side comparisons

```{r}
commonletters
nos_letters
```

```{r}
commonf3
nos_f3
```

```{r}
commonm3
nos_m3
```

```{r}
commone3
nos_e3
```

Some points of my wordle strategy:

- Start with the same word every time
- Use 5 different letters
- If any letters match, I use them in the next guess
- Do not aim to get the word on the first guess

Takeaway points so far:

- S is the most common first and last letter 
- ED is the most common ending if you remove S from the 5th letter
- A can be found in the middle 3 letters all the time

If I were to follow the above, then words like SARED, SATED, SAWED, SAPED, SAKER, SANNY, SAPPY would all be potential first guesses.
However, these words will only be the correct *once* over the span of many years. So if I dont have any matches in my first guess, I should have a second one ready that covers the most common letters

Let's say we started with SATED, and ended up getting no matches

```{r}
nosated <- wordlist
nosated$word <- wordlist$word %>% str_remove_all("[SATED]{1,}") 

nosated <- nosated %>% 
  filter(str_detect(nosated$word, "^.{5}"))

letterplot <- function(worddf, columns){
  worddf %>% 
  select(contains(columns)) %>% 
    pivot_longer(everything(), names_to = "pos", values_to = "letter") %>% 
    group_by(pos) %>% 
    count(letter) %>% 
    arrange(desc(n), .by_group = TRUE) %>% 
    slice_head(n = 5) %>% 
    ggplot(aes(x = fct_reorder(letter, -n), y = n, fill = pos))+
    geom_col(position = position_dodge())
}

letterplot(nosated, "L")+
  labs(x = "Letter",
       y = "",
       title = "Most common letter positions",
       subtitle = "After A E D T S")
```

The letters O I N R L all appear multiple times, unfortunately there isn't a word on the list that I can make with these letters

Going back, our first word was SATED which leaves our ideal opening words to contain:

- AEIO for vowels
- NRLTDS for consonants

With some guess, check and revise, there are probably many 2 word combinations from these 10 letters.
Now let's say after using those 10 letters you still didnt have any matches

```{r}
uncommons <- wordlist

uncommons$word <- uncommons$word %>% 
  str_remove_all("[AEIONRLTDS]")

uncommons <- uncommons %>% filter(str_detect(uncommons$word, "^.{5}"))

uncommons %>% select(word)
```

This leaves only 35 possible words. So it's safe to say that the letters AEIONRLTDS covers a vast majority of English words so your first guess in Wordle should include these

*For the Scrabble players: These letters are all worth 1 point in Scrabble*

Concluding notes:

- This doesn't mean that you should always start off with things like ADORE and LINTS for your first 2 guesses. 
- The only thing I can guarantee is that it's a high probability that the word has at least *one* of these 10 letters
- An example would be the word I used at the time of writing this (3/23/22) which is PURGE. Only 2 of the top 10 letters are used 

